The Galactic Midplane is Not a Plane: Implications for {\em Gaia}

### 2019-04-22 paper/ms.tex
completed initial revisions to introduction

Once a Galactocentric coordinate system has been established and a 6D phase
space measurement of a star has been made, it is often desirable to convert
this measurement into action space to concisely summarize its projected orbit,
model the stellar distribution function, or find stars with similar dynamical
properties. Actions are conserved quantities that describe the orbit of a
star under the assumption of regular, bound orbits in a system where the
equations of motion are separable in a particular coordinate system. They are
the %mm typical
 integral of the Hamiltonian in cylindrical coordinates and are
given by,
\beq\label{eq:actions}
J_i \equiv
\frac{1}{2\pi} \oint_{\text{orbit}}p_i\,\text{d}x_i\text{,}
\eeq
where $i=R,z,\phi$ are the radial, vertical, and azimuthal coordinates
respectively in a cylindrical coordinate system and the $p_i$ are the
conjugate momenta.\footnote{We typically write $J_r$ instead of $J_R$ for
aesthetic reasons.} In a slowly-evolving axisymmetric potential, these actions
are invariant and $J_{\phi} \equiv L_z$
\citep{2008gady.book.....B,2014RvMP...86....1S}.

With the advent of 6D phase-space measurements over a relatively large
($\gtrsim 2$ kpc) volume from the {\em Gaia} satellite, the study of stars'
actions has gained new popularity. One reason is dimensionality reduction---an
individual stellar orbit is concisely described by three actions, as opposed
to six dimensions of phase space. Second, under the assumption of a
phase-mixed system, the dynamical properties of a population of stars should
be uniquely a function of their actions and independent of the conjugate
angles, allowing investigation of the relationship between {\em orbital}
properties of stars and other intrinsic, and at least partially invariant,
properties such as age or metallicity. Actions also provide a convenient basis
for constructing models of the stellar distribution function, or for
associating stars with similar dynamical properties (e.g. to determine
membership in moving groups).

@@

The transformation to action space is nonlinear in both the phase-space
position and the potential, and requires relatively strong assumptions about
the system's symmetry. For instance, the Galactic potential is not strictly
axisymmetric; with the vast improvement in the quality of phase-space
measurements due to {\em Gaia}, this assumption appears increasingly
inadequate \citep[e.g.][]{2018Natur.561..360A}. Even if this assumption is
close enough for many purposes, the parameters used in axisymmetric models may
be inadequately constrained by current observations.

In this work we discuss how the imposition of a global axisymmetric coordinate
system based on a local measurement of the midplane location and
Galactocentric distance introduces systematic error in the computation of
actions, especially for stars in the thin disk. Our main point is that the
``local'' Galactic midplane is not the same as the ``global'' Galactic
midplane. For stars relatively far from the Sun, the local midplane location
will not agree with our local midplane extrapolated onto their position,
thanks to relatively small-scale variations in the density of the gas and
stellar distribution. Converting positions and velocities of more distant
stars from a heliocentric to a Galactocentric coordinate system introduces a
systematic bias in the $z$ coordinate, and hence the actions computed for
these stars. The further from the solar neighborhood the target star, the more
likely the mismatch will result in large systematic uncertainty, especially in
the vertical action $J_z$. Depending on the type of orbit being studied, this
sets a limit on the ``dynamical solar neighborhood'' over which the local
coordinate frame can be usefully extended for action-space analysis. A similar
argument applies to uncertainty in measurements of $R_{\odot}$\footnote{Here,
and throughout, we use $R_{\odot}$ to mean the distance from the Sun to the
Galactic center.} and variations in the local standard of rest (LSR).

In Section~\ref{sec:ref_frame}, we describe the general impact coordinate
system errors have on the measured actions. Lacking an empirical, nonlocal
definition of the global Galactic midplane or a measurement of the expected
midplane variation in the Milky Way itself,  in Section~\ref{sec:local_fire}
we attempt to quantify the azimuthal dependence of the local midplane, and
hence the expected size of the dynamical solar neighborhood, using
cosmological-hydrodynamical zoom-in simulations of Milky Way-mass galaxies
from the Feedback in Realistic Environments (FIRE)
collaboration\footnote{\url{https://fire.northwestern.edu}}
\citep{2016ApJ...827L..23W,2017arXiv170206148H}.

In Section~\ref{sec:discussion} we discuss the implications of midplane
variations, and the resulting systematic uncertainty in the vertical action,
for action-space analyses beyond the dynamical solar neighborhood. As an
illustration, in Section \ref{sssec:neighborhood} we use these estimates to
consider the feasibility of one science goal that relies on the invariance of
actions --- the reconstruction of open clusters across large distances using
``dynamical tagging;'' that is, the association of stars to a birth cluster
based on their similar actions. In light of our findings, we considered the
orbits of the nine open clusters within $250\,\pc$ as reported by the {\em
Gaia} collaboration \citep{2018AA...616A..10G}. In
Table~\ref{tab:real_clusters}, we report the distance to each cluster, the
three actions, and the maximum vertical extent of their orbits
($z_{\text{max}}$). In Section~\ref{sssec:reconstruction} we estimate and
discuss the size of the dynamical solar neighborhood for each cluster based on
its mass, to show that action-space tagging is likely still a viable method if
contamination from foreground or background stars is unimportant. We summarize
and conclude in Section~\ref{sec:conclusion}.

>>>>

Once a Galactocentric coordinate system has been established and a
six-dimensional (6D) phase
space measurement of a star has been made, it is often desirable to convert
this measurement into action space to concisely summarize its projected orbit,
model the stellar distribution function, or find stars with similar dynamical
properties. Actions are conserved quantities that describe the orbit of a
star under the assumption of regular, bound orbits in a system where the
equations of motion are separable in a particular coordinate system. They are
the 
%mm typical
integral of the Hamiltonian in cylindrical coordinates 
%mm and are given by,
\beq\label{eq:actions}
J_i \equiv
\frac{1}{2\pi} \oint_{\text{orbit}}p_i\,\text{d}x_i\text{,}
\eeq
where $i=R,z,\phi$ are the radial, vertical, and azimuthal coordinates
respectively in a cylindrical coordinate system and the $p_i$ are the
conjugate momenta.\footnote{We 
%mm typically [there are no other J_R's in manuscript]
write $J_r$ instead of $J_R$ for
aesthetic reasons.} In a slowly-evolving axisymmetric potential, these actions
are invariant and $J_{\phi} \equiv L_z$
\citep{2008gady.book.....B,2014RvMP...86....1S}.

With the advent of 6D phase-space measurements over a relatively large
($\gtrsim 2$ kpc) volume from the {\em Gaia} satellite, the study of stars'
actions has gained new popularity. One reason is dimensionality reduction---an
individual stellar orbit is concisely described by three actions, as opposed
to six phase space coordinates. Second, under the assumption of a
phase-mixed system, the dynamical properties of a population of stars should
be uniquely a function of their actions and independent of the conjugate
angles, allowing investigation of the relationship between {\em orbital}
properties of stars and other intrinsic, and at least partially invariant,
properties such as age or metallicity. Actions also provide a convenient basis
for constructing models of the stellar distribution function, or for
associating stars with similar dynamical properties (e.g. to determine
membership in moving groups).

@@

The transformation to action space is nonlinear in both the phase-space
position and the potential, and requires relatively strong assumptions about
the system's symmetry. 
%mm For instance, 
    However, 
the Galactic potential is not strictly
axisymmetric; with the vast improvement in the quality of phase-space
measurements due to {\em Gaia}, this assumption appears increasingly
inadequate \citep[e.g.][]{2018Natur.561..360A}. Even if this
assumption 
%mm is [counterfactual subjunctive, I believe]
     were
close enough for many purposes, the parameters used in axisymmetric models may
be inadequately constrained by current observations.

In this work we discuss how the imposition of a global axisymmetric coordinate
system based on a local measurement of the midplane location and
Galactocentric distance introduces systematic error in the computation of
actions, especially for stars in the thin disk. Our main point is that the
``local'' Galactic midplane is not the same as the ``global'' Galactic
midplane. For stars relatively far from the Sun, the local midplane location
will not agree with our local midplane extrapolated onto their position,
thanks to relatively small-scale variations in the density of the gas and
stellar distribution. Converting positions and velocities of more distant
stars from a heliocentric to a Galactocentric coordinate system introduces a
systematic bias in the $z$ coordinate, and hence the actions computed for
these stars. The further from the solar neighborhood the target star, the more
likely the mismatch will result in large systematic uncertainty, especially in
the vertical action $J_z$. Depending on the type of orbit being studied, this
sets a limit on the ``dynamical solar neighborhood'' over which the local
coordinate frame can be usefully extended for action-space analysis. A similar
argument applies to 
%mm [given that this is down to 0.3% error, seems a bit academic]
   any remaining 
uncertainty in measurements of 
%mm $R_{\odot}$\footnote{Here,
%and throughout, we use $R_{\odot}$ to mean the distance from the Sun to the
%Galactic center.}
     the Galactocentric radius of the Sun $R_{\odot}$
and to variations in the local standard of rest (LSR).

In Section~\ref{sec:ref_frame}, we describe the general impact coordinate
system errors have on the measured actions. Lacking an empirical, nonlocal
definition of the global Galactic midplane or a measurement of the expected
midplane variation in the Milky Way itself,  in Section~\ref{sec:local_fire}
we 
%mm attempt to quantify the 
    examine the expected 
azimuthal dependence of the local midplane, and
hence the expected size of the dynamical solar neighborhood, using
cosmological, hydrodynamical, zoom-in simulations of Milky Way-mass galaxies
from the Feedback in Realistic Environments (FIRE)
collaboration\footnote{\url{https://fire.northwestern.edu}}
\citep{2016ApJ...827L..23W,2017arXiv170206148H}.

In Section~\ref{sec:discussion} we discuss the implications of midplane
variations, and the resulting systematic uncertainty in the vertical action,
for action-space analyses beyond the dynamical solar neighborhood. As an
illustration, in Section \ref{sssec:neighborhood} we use these estimates to
consider the feasibility of one science goal that relies on the invariance of
actions---the reconstruction of open clusters across large distances using
``dynamical tagging;'' that is, the association of stars to a birth cluster
based on their similar actions. In light of our findings, we considered the
orbits of the nine open clusters within $250\,\pc$ as reported by the {\em
Gaia} collaboration \citep{2018AA...616A..10G}. In
Table~\ref{tab:real_clusters}, we report the distance to each cluster, the
three actions, and the maximum vertical extent of their orbits
($z_{\text{max}}$). In Section~\ref{sssec:reconstruction} we estimate and
discuss the size of the dynamical solar neighborhood for each cluster based on
its mass, to show that action-space tagging is likely still a viable method if
contamination from foreground or background stars is unimportant. We summarize
and conclude in Section~\ref{sec:conclusion}.
